\Large\textbf{PixelBytes: Catching Unified Representation for Multimodal Generation}

### 2024-09-11 docs/paper/article.tex
update figure

% Title page
\title{\Large\textbf{PixelBytes: Catching Unified Embedding for Multimodal Generation}}
\author{\large Fabien Furfaro}
\date{\large 2024}

@@

\begin{abstract}
This report introduces PixelBytes Embedding, a novel approach for unified multimodal representation learning. Our method captures diverse inputs in a single, cohesive representation, enabling emergent properties for multimodal sequence generation, particularly for text and pixelated images. Inspired by state-of-the-art sequence models such as Image Transformers, PixelCNN, and Mamba-Bytes, PixelBytes aims to address the challenges of integrating different data types. We explore various model architectures, including Recurrent Neural Networks (RNNs), State Space Models (SSMs), and Attention-based models, focusing on bidirectional processing and our innovative PxBy embedding technique. Our experiments, conducted on a specialized PixelBytes Pokémon dataset, demonstrate that bidirectional sequence models with PxBy embedding and convolutional layers can generate coherent multimodal sequences. This work contributes to the advancement of integrated AI models capable of understanding and generating multimodal data in a unified manner. Code is available at \url{https://github.com/fabienfrfr}.
\end{abstract}

\let\thefootnote\relax\footnotetext{$^{*}$ Corresponding author: Fabien Furfaro (\url{fabien.furfaro@gmail.com}).}

\begin{figure}[h]
\centering
\includegraphics[width=0.8\linewidth]{example_generation.png}
\caption{Example of a PixelBytes-generated Pokémon and its description. Left: Generated pixelated image. Right: Corresponding text description. The red box indicates the model's attention window for both pixels and words. Note that the model currently exhibits some inaccuracies in image generation and occasionally invents words, highlighting areas for future improvement.}
\label{fig:example_generation}
\end{figure}

@@

Recent advancements in artificial intelligence have led to increasingly generalist models, not by combining multiple specialized components, but by giving simple tasks to models where emergent properties—complex behaviors that arise from simpler underlying rules—appear. This is the case with generative language models like GPT \cite{brown2020language}. However, these models are limited by their focus on language alone, failing to capture the full complexity of multimodal understanding \cite{alayrac2023language}. To address this limitation, researchers have explored combining LLMs with other modalities \cite{radford2021learning, ramesh2022hierarchical}. But this brings us back to the initial problem, as it often results in specialized model combinations without allowing for new emergent properties. We propose "PixelBytes Embedding", a novel approach enabling unified training across modalities by capturing diverse inputs in a single, cohesive representation.

Multimodal sequence generation, which involves the creation of coherent outputs combining various data types such as text, images, and numerical sequences, presents a significant challenge in artificial intelligence \cite{baltrusaitis2019multimodal}. While models like GPT have excelled in text generation \cite{brown2020language}, there's a growing need for unified approaches that can seamlessly handle varied data types. Building upon these findings, PixelBytes aims to address the challenge of unified text and image generation by proposing a model capable of producing mixed sequences of text and images in a coherent and unified manner. We draw inspiration from state-of-the-art sequence models, including Image Transformers \cite{parmar2018image}, PixelCNN \cite{oord2016conditional}, and recent developments in bytes generation by mamba architectures \cite{wang2024mamba}.

Our research explores various architectures including Recurrent Neural Networks (RNNs), State Space Models (SSMs), and Attention-based models, focusing on the effectiveness of bidirectional processing \cite{gu2023visual, gu2023bidirectional}, novel embedding techniques (particularly PxBy embedding, which unifies pixel and byte-level representations), the impact of convolutional layers, effects of model depth, input dimensionality, and size. Our experiments reveal that bidirectional sequence models with PxBy embedding and convolutional layers demonstrate good performance in generating coherent sequences.

Our results show that it is possible to train sequence models with unified embeddings for both text and pixel completion. This paper presents our methodology to construct unified sequence data from text and image, experimental results, and analysis, contributing to the development of our sequence generation embedding models. The proposed PixelBytes Embedding approach has the potential to significantly advance the field of multimodal AI, enabling more integrated and versatile models capable of understanding and generating diverse types of data. This could lead to improvements in areas such as visual question answering, image captioning, and multimodal content generation for virtual and augmented reality applications.

@@

\subsection{Overview}
The PixelBytes architecture is designed to seamlessly integrate multimodal data for unified sequence generation. At its core, the model incorporates two key innovative components: a specialized tokenizer sequence constructor and a unified multimodal embedding technique called PxByEmbed. The tokenizer sequence constructor is engineered to process both pixelated images and text at a byte level, enabling a consistent representation across modalities. This is complemented by PxByEmbed, our novel embedding approach that creates a unified representation for both pixel and byte data in a single, coherent space. This architecture draws inspiration from recent advancements in multimodal learning \cite{baltrusaitis2019multimodal} and efficient sequence modeling \cite{vaswani2017attention}, allowing PixelBytes to capture intrinsic relationships between visual and textual information effectively.

@@

Image captioning datasets, while combining visual and textual modalities, prove unsuitable for joint text and image generation due to limited text content and difficulties in interpreting pixelated versions of high-resolution images. To address these limitations, we developed a specialized Pokémon dataset, offering advantages such as a long-standing franchise history, pixelated designs, rich descriptive text, and over 1000 unique Pokémon. This aligns with trends in creating specialized datasets for multimodal learning \cite{lin2014microsoft}.

We constructed our dataset by web scraping Pokémon miniatures and descriptions from Pokepedia using Beautiful Soup \cite{richardson2007beautiful}, maintaining a 2/3 text to 1/3 image ratio. For image quantization, we employed a 55-color palette inspired by the NES, creating tokens representing different color combinations. This approach translates visual information into a format suitable for sequence modeling, similar to recent work in discrete representation learning for images \cite{van2017neural}. The quantization and pixelization process used OpenCV and scikit-image \cite{bradski2000opencv, vanderwalt2014scikit}. To ensure balanced representation, we adjusted the number of image and text tokens to 113 characters for each modality, allowing the model to learn equally from both visual and textual information. The final dataset, balancing text and pixelated images, is available on the Hugging Face Datasets Hub for reproducibility at \url{https://huggingface.co/datasets/ffurfaro/PixelBytes-Pokemon}.

@@

\begin{algorithm}[h]
\caption{PxByEmbed: Multimodal Embedding Algorithm (k=3)
\newline
\textbf{Input:} $V$: vocabulary size, $D$: embedding dimension
\newline
\textbf{Output:} Embedded representation $\mathbf{E} \in \mathbb{R}^{B \times L \times D}$
\newline
\textbf{Note:} $\mathbf{X}_{emb} \in \mathbb{R}^{B \cdot L \times E_{int} \times k \times k}$, 
$\mathbf{X}_{flat} \in \mathbb{R}^{B \cdot L \times E_{int}k^2}$, 
$\mathbf{X}_{proj} \in \mathbb{R}^{B \cdot L \times D}$
}
\begin{algorithmic}[0]
\State \textbf{Initialize:}
\State $k \gets 3$
\State $E_{int} \gets \max(9, \lfloor D / k^2 \rfloor)$
\State $\mathbf{\alpha} \in \mathbb{R}^{1 \times 1 \times k \times k}$ 
\State $\mathbf{W}_{emb} \in \mathbb{R}^{V \times E_{int}}$ 
\State $\mathbf{W}_{proj} \in \mathbb{R}^{E_{int}k^2 \times D}$ 
\State $\mathbf{W}_{patch} \in \mathbb{R}^{E_{int} \times E_{int} \times k \times k}$

@@

PixelBytes uses a method to handle transitions between text and image tokens during sequence generation. This approach is used in both dataset construction and sequence generation (Figure \ref{fig:example_generation}).

@@

\subsection{Model Architectures}
We evaluated three compact model architectures, each with fewer than 100,000 parameters: a Recurrent Neural Network (RNN) using bidirectional Long Short-Term Memory (LSTM) units \cite{hochreiter1997long}, a Transformer \cite{vaswani2017attention}, and Mamba, based on the State Space Model (SSM) \cite{gu2022efficiently}. These models were adapted to process our dataset of pixel data and bytecode sequences. Training was conducted on Kaggle using dual T4 GPUs, with a batch size of 32, sequence length of 256, stride size of 32, and learning rate of 0.001. We trained for 200 epochs, evaluating performance every 5 epochs. The resulting models are available at \url{https://huggingface.co/ffurfaro/PixelBytes-Pokemon}. Additionally, we developed a more specialized model for generation tasks, which is not presented in this paper.

@@

Our experiments with various model architectures tested the effectiveness of unified text and image generation. We evaluated model performance using Hamming distance, cosine similarity, and BLEU score \cite{papineni2002bleu}. Bidirectional RNN models using PixelBytes (PxBy) embedding with convolutional layers generally performed better across our metrics. The PxBy embedding improved the mean BLEU score from 0.750 to 0.777 compared to a center-only approach, indicating its effectiveness in representing multimodal data. This finding aligns with recent work emphasizing the importance of effective multimodal embeddings in text-to-image generation tasks \cite{yan2024multimodal}.

Bidirectional models showed a slight advantage over unidirectional ones, with BLEU scores of 0.777 and 0.776 respectively. This suggests that bidirectionality may help capture sequential dependencies, consistent with other studies in sequence modeling \cite{schuster1997bidirectional}. Notably, varying state dimensions (32, 64, 128) had minimal impact on performance, with BLEU scores ranging from 0.775 to 0.784. While RNN models showed strong performance in our experiments, State Space Models (SSM) demonstrated rapid convergence during training, suggesting potential for future exploration in multimodal data processing. This is in line with recent advancements in SSM for sequence modeling tasks \cite{gu2022efficiently}.

@@

This study on PixelBytes highlights the potential for unified text-image generation, contributing to the growing field of multimodal AI \cite{baltrusaitis2019multimodal}. The performance of bidirectional RNN models with PxBy embedding underscores the importance of capturing both spatial and contextual information in multimodal data. Future work could focus on refining the PxByEmbed algorithm, testing on larger and more diverse datasets, exploring creative applications of multimodal generation, and further optimizing SSM and Attention models. These efforts aim to advance multimodal sequence modeling and generation techniques, addressing challenges identified in surveys of the field \cite{lin2014microsoft}.

>>>>

% Title page
\title{\Large\textbf{PixelBytes: Catching Unified Representation for Multimodal Generation}}
\author{\large Fabien Furfaro $^{*}$}
\date{\large 2024}

@@

\begin{abstract}
This report introduces PixelBytes, a novel approach for unified multimodal representation learning. Our method captures diverse inputs in a single, cohesive representation, enabling emergent properties for multimodal sequence generation, particularly for text and pixelated images. Inspired by state-of-the-art sequence models such as Image Transformers, PixelCNN, and Mamba-Bytes, PixelBytes aims to address the challenges of integrating different data types. We explore various model architectures, including Recurrent Neural Networks (RNNs), State Space Models (SSMs), and Attention-based models, focusing on bidirectional processing and our innovative PxBy embedding technique. Our experiments, conducted on a specialized PixelBytes Pokémon dataset, demonstrate that bidirectional sequence models with PxBy embedding and convolutional layers can generate coherent multimodal sequences. This work contributes to the advancement of integrated AI models capable of understanding and generating multimodal data in a unified manner. Code is available at \url{https://github.com/fabienfrfr/PixelBytes}.
\end{abstract}

\let\thefootnote\relax\footnotetext{$^{*}$ Corresponding author : Fabien Furfaro (\url{fabien.furfaro@gmail.com}).}

\begin{figure}[h]
\centering
\includegraphics[width=0.8\linewidth]{approach_generation.png}
\caption{PixelBytes Pokémon generation process. Left: Image and text tokenization with embedding window for modality unification. Center: Sequence generation using PxBy Embedding. Right: Generated pixelated image and text description. Red box shows a possible generation example. Note: The model currently has some inaccuracies in image generation and may invent words, indicating areas for future improvement.}
\label{fig:approach_generation}
\end{figure}

Amélioration de l'idée : il y a trop de sequence qui se repete ! finaleemnt, la proposition de l'embedding n'est pas bonne ! les resultats avec plus d'approfondissement ne montre pas vraiment que l'embedding avec un convolution est efficace. plutot utiliser les 6 positions specifique (dim entree = dim sortie) et agrandire l'embedding, la génération est mieux (voir Figure 1) ! de meme simplification du code necessaire ! le tokenizer doit etre plus flexible, si on a besoin de générer les donner en plus, ca devient compliqué ! il faut que le tokenizer fasse tout ! (ou utiliser l'egalité entre la dimension de sortie et de l'entrée pour faire un vrai modele autoregressif) Donc dans l'article, il faut recommancer, il y aura donc deux partie ! d'abords la premiere version, pour exploration --> on voit que l'embedding n'est pas suffisant, donc on passe à la nouvelle approche ! (ne pas oublier de comparer avec ImageBind, l'idée n'est pas d'avoir des intermediaire qui ne permette pas l'emergence de propriété)

L'idée n'est pas de faire un modele "ultra performant", sinon pour ces taches, les modeles de diffsion sont mieux (Why Does Diffusion Work Better than Auto-Regression?) maais plutot des mdoele generaux de "fondation" qui ont appris toutes les modalités d'un coup. Ici on ne test que les modele auto-regressif, mais des approche comme Diffusion-LM pourrait etre une autre proposition (et compatible avec notre tokenizer)

@@

Recent advancements in artificial intelligence have led to increasingly generalist models, not by combining multiple specialized components, but by giving simple tasks to models where emergent properties—complex behaviors that arise from simpler underlying rules—appear. This is the case with generative language models like GPT \cite{brown2020language}. However, these models are limited by their focus on language alone, failing to capture the full complexity of multimodal understanding \cite{huang2023language}. To address this limitation, researchers have explored combining LLMs with other modalities \cite{liu2024visual}. But this brings us back to the initial problem, as it often results in specialized model combinations without allowing for new emergent properties. We propose "PixelBytes Embedding", a novel approach enabling unified training across modalities by capturing diverse inputs in a single, cohesive representation.

Multimodal sequence generation, which involves the creation of coherent outputs combining various data types such as text, images, and numerical sequences, presents a significant challenge in artificial intelligence \cite{baltruvsaitis2018multimodal}. While models like GPT have excelled in text generation \cite{brown2020language}, there's a growing need for unified approaches that can seamlessly handle varied data types. Building upon these findings, PixelBytes aims to address the challenge of unified text and image generation by proposing a model capable of producing mixed sequences of text and images in a coherent and unified manner. We draw inspiration from state-of-the-art sequence models, including Image Transformers \cite{parmar2018image}, PixelCNN \cite{van2016conditional}, and recent developments in bytes generation by mamba architectures \cite{wang2024mambabyte}.

Our research explores various architectures including Recurrent Neural Networks (RNNs), State Space Models (SSMs) \cite{dao2024transformers}, and Attention-based models, focusing on the effectiveness of bidirectional processing \cite{zhu2024vision}, novel embedding techniques (particularly PxBy embedding, which unifies pixel and byte-level representations), the impact of convolutional layers, effects of model depth, input dimensionality, and size. Our experiments reveal that bidirectional sequence models with PxBy embedding and convolutional layers demonstrate good performance in generating coherent sequences.

Our results show that it is possible to train sequence models with unified embeddings for both text and pixel completion. This paper presents our methodology to construct unified sequence data from text and image, experimental results, and analysis, contributing to the development of our sequence generation embedding models. The proposed PixelBytes Embedding approach may offer valuable insights for the development of multimodal AI, potentially leading to models that can better understand and generate various types of data.

@@

\subsection{Overview}
The PixelBytes architecture is designed to seamlessly integrate multimodal data for unified sequence generation. At its core, the model incorporates two key innovative components: a specialized tokenizer sequence constructor and a unified multimodal embedding technique called PxByEmbed. The tokenizer sequence constructor is engineered to process both pixelated images and text at a byte level, enabling a consistent representation across modalities. This is complemented by PxByEmbed, our novel embedding approach that creates a unified representation for both pixel and byte data in a single, coherent space. This architecture draws inspiration from recent advancements in multimodal learning \cite{baltruvsaitis2018multimodal} and efficient sequence modeling \cite{vaswani2017attention}, allowing PixelBytes to capture intrinsic relationships between visual and textual information effectively.

@@

Image captioning datasets, while combining visual and textual modalities, prove unsuitable for joint text and image generation due to limited text content and difficulties in interpreting pixelated versions of high-resolution images. To address these limitations, we developed a specialized Pokémon dataset, offering advantages such as a long-standing franchise history, pixelated designs, rich descriptive text, and over 1000 unique Pokémon.

We constructed our dataset by web scraping Pokémon miniatures and descriptions from Pokepedia using Beautiful Soup \cite{richardson2007beautiful}, maintaining a 2/3 text to 1/3 image ratio. For image quantization, we employed a 55-color palette inspired by the NES, creating tokens representing different color combinations. This approach translates visual information into a format suitable for sequence modeling. The quantization and pixelization process used OpenCV and scikit-image \cite{bradski2000opencv, van2014scikit}. To ensure balanced representation, we adjusted the number of image and text tokens to 113 indices for all entries, allowing the model to learn equally from both visual and textual information. The final dataset, balancing text and pixelated images, is available on the Hugging Face Datasets Hub for reproducibility at \url{https://huggingface.co/datasets/ffurfaro/PixelBytes-Pokemon}.

@@

\begin{algorithm}[ht]
\caption{PxByEmbed: Multimodal Embedding Algorithm (k=3)
\newline
\textbf{Input:} $V$: vocabulary size, $D$: embedding dimension
\newline
\textbf{Output:} Embedded representation $\mathbf{E} \in \mathbb{R}^{B \times L \times D}$
\newline
\textbf{Note:} $\mathbf{X}_{emb} \in \mathbb{R}^{B \cdot L \times E_{int} \times k \times k}$, 
$\mathbf{X}_{flat} \in \mathbb{R}^{B \cdot L \times E_{int}k^2}$, 
$\mathbf{X}_{proj} \in \mathbb{R}^{B \cdot L \times D}$
}
\begin{algorithmic}[0]
\State \textbf{Initialize:}
\State $k \gets 3$
\State $E_{int} \gets \max(9, \lfloor D / k^2 \rfloor)$
\State $\mathbf{\alpha} \in \mathbb{R}^{1 \times 1 \times k \times k}$ 
\State $\mathbf{W}_{emb} \in \mathbb{R}^{V \times E_{int}}$ 
\State $\mathbf{W}_{proj} \in \mathbb{R}^{E_{int}k^2 \times D}$ 
\State $\mathbf{W}_{patch} \in \mathbb{R}^{E_{int} \times E_{int} \times k \times k}$

@@

PixelBytes uses a method to handle transitions between text and image tokens during sequence generation. This approach is used in both dataset construction and sequence generation (Figure \ref{fig:approach_generation}).

@@

\subsection{Model Architectures}
We evaluated three compact model architectures, each with fewer than 100,000 parameters: a Recurrent Neural Network (RNN) using bidirectional Long Short-Term Memory (LSTM) units \cite{schmidhuber1997long}, a Transformer \cite{vaswani2017attention}, and Mamba, based on the State Space Model (SSM) \cite{dao2024transformers}. These models were adapted to process our dataset of pixel data and bytecode sequences. Training was conducted on Kaggle using dual T4 GPUs, with a batch size of 32, sequence length of 256, stride size of 32, and learning rate of 0.001. We trained for 200 epochs, evaluating performance every 5 epochs. The resulting models are available at \url{https://huggingface.co/ffurfaro/PixelBytes-Pokemon}. Additionally, we developed a more specialized model for generation tasks, which is not presented in this paper.

@@

Our experiments with various model architectures tested the effectiveness of unified text and image generation. Bidirectional RNN models using PixelBytes (PxBy) embedding with convolutional layers generally performed better across our metrics. The PxBy embedding improved the mean BLEU score from 0.750 to 0.777 compared to a center-only approach, indicating its effectiveness in representing multimodal data. This finding aligns with recent work emphasizing the importance of effective multimodal embeddings in text-to-image generation tasks \cite{verHo2021efficient}.

Bidirectional models showed a slight advantage over unidirectional ones, with BLEU scores of 0.777 and 0.776 respectively. This suggests that bidirectionality may help capture sequential dependencies, consistent with other studies in sequence modeling \cite{schuster1997bidirectional}. Notably, varying state dimensions (32, 64, 128) had minimal impact on performance, with BLEU scores ranging from 0.775 to 0.784. While RNN models showed strong performance in our experiments, State Space Models (SSM) demonstrated rapid convergence during training, suggesting potential for future exploration in multimodal data processing. This is in line with recent advancements in SSM for sequence modeling tasks \cite{dao2024transformers}.

@@

This study on PixelBytes highlights the potential for unified text-image generation, contributing to the growing field of multimodal AI \cite{baltruvsaitis2018multimodal}. The performance of bidirectional RNN models with PxBy embedding underscores the importance of capturing both spatial and contextual information in multimodal data. Future work could focus on refining the PxByEmbed algorithm, testing on larger and more diverse datasets, exploring creative applications of multimodal generation, and further optimizing SSM and Attention models. These efforts aim to advance multimodal sequence modeling and generation techniques.
